---
title: |-
  Bradley-Terry Models in [R]{.sans-serif}: The
  **BradleyTerry2** Package\
  For **BradleyTerry2** version
  `r packageDescription("BradleyTerry2")[["Version"]]`, `r Sys.Date()`\
  <https://github.com/hturner/BradleyTerry2>\
abstract: |
  This is a short overview of the R add-on package **BradleyTerry2**,
  which facilitates the specification and fitting of Bradley-Terry
  logit, probit or cauchit models to pair-comparison data. Included are
  the standard 'unstructured' Bradley-Terry model, structured versions
  in which the parameters are related through a linear predictor to
  explanatory variables, and the possibility of an order or 'home
  advantage' effect or other 'contest-specific' effects. Model fitting
  is either by maximum likelihood, by penalized quasi-likelihood (for
  models which involve a random effect), or by bias-reduced maximum
  likelihood in which the first-order asymptotic bias of parameter
  estimates is eliminated. Also provided are a simple and efficient
  approach to handling missing covariate data, and suitably-defined
  residuals for diagnostic checking of the linear predictor.
author: |
  Heather Turner\
  University of Warwick\
  David Firth\
  University of Warwick
date: '2024-08-14'
vignette: |-
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Bradley-Terry models in R}
  %\VignetteDepends{}
output:
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    number_sections: no
    includes:
      in_header: auto-number-sec-js.html
link-citations: yes
bibliography: BradleyTerry-generated.bib

---

::: article
``` {r include=FALSE}
library(knitr)
opts_chunk$set(
tidy=FALSE
)
```

@

``` {r set_options, echo = FALSE}
options(prompt = "R> ", continue = "+  ", width = 70,
        useFancyQuotes = FALSE, digits = 7)
```

# Introduction

The Bradley-Terry model [@brad:terr:52] assumes that in a 'contest'
between any two 'players', say player $i$ and player $j$
$(i, j \in \{1,\ldots,K\})$, the odds that $i$ beats $j$ are
$\alpha_i/\alpha_j$, where $\alpha_i$ and $\alpha_j$ are positive-valued
parameters which might be thought of as representing 'ability'. A
general introduction can be found in @brad:84 or @agre:02. Applications
are many, ranging from experimental psychology to the analysis of sports
tournaments to genetics (for example, the allelic
transmission/disequilibrium test of [@sham:curt:95] is based on a
Bradley-Terry model in which the 'players' are alleles). In typical
psychometric applications the 'contests' are comparisons, made by
different human subjects, between pairs of items.

:::: {#algo1}
```{.algorithm}
% This quicksort algorithm is extracted from Chapter 7, Introduction to Algorithms (3rd edition)
\begin{algorithm}
\caption{Quicksort}
\begin{algorithmic}
\PROCEDURE{Quicksort}{$A, p, r$}
    \IF{$p < r$} 
        \STATE $q = $ \CALL{Partition}{$A, p, r$}
        \STATE \CALL{Quicksort}{$A, p, q - 1$}
        \STATE \CALL{Quicksort}{$A, q + 1, r$}
    \ENDIF
\ENDPROCEDURE
\PROCEDURE{Partition}{$A, p, r$}
    \STATE $x = A[r]$
    \STATE $i = p - 1$
    \FOR{$j = p$ \TO $r - 1$}
        \IF{$A[j] < x$}
            \STATE $i = i + 1$
            \STATE exchange
            $A[i]$ with $A[j]$
        \ENDIF
        \STATE exchange $A[i]$ with $A[r]$
    \ENDFOR
\ENDPROCEDURE
\end{algorithmic}
\end{algorithm}
```
::::

::::{#master_algo}
```{.algorithm}
\begin{algorithm}
	\caption{Algorithm for the maximization of a continuously differentiable multi-convex function}
	\begin{algorithmic}
		\STATE {\bfseries Result:} {$\mathbf a_1^s, \ldots,  \mathbf a_J^s$ (approximate solution of ($\ref{eq:unstructured}$))}
		\STATE {\bfseries Initialization:} {choose random vector $\mathbf a_j^0\in\Omega_j, j =1, \ldots, J$, $\varepsilon$;}
		\STATE$s = 0$ ;
		\REPEAT
		\FOR{$j=1$ {\bfseries to} $J$}
		\STATE $\vcenter{\begin{equation}
		     \mathbf a_j^{s+1} = r_j\left(  \mathbf a_1^{s+1}, \ldots,  \mathbf a_{j-1}^{s+1},  \mathbf a_j^{s}, \ldots,  \mathbf a_J^{s}\right).
		\end{equation}}$
		\ENDFOR
		\STATE$s = s + 1$ ;
		\UNTIL{$f( \mathbf a_1^{s+1}, \ldots,  \mathbf a_J^{s+1})-f( \mathbf a_1^s, \ldots,  \mathbf a_J^s) < \varepsilon$}
	\end{algorithmic}
\end{algorithm}
```
::::

ref to [algo1](#algo1)

or ref like this: \@ref(algo1)

The model can alternatively be expressed in the logit-linear form
$$\mathop{\rm logit}[\mathop{\rm pr}(i\ \mathrm{beats}\ j)]=\lambda_i-\lambda_j,
\label{eq:unstructured}   (\#eq:unstructured)$$
where $\lambda_i=\log\alpha_i$ for all $i$. Thus, assuming independence
of all contests, the parameters $\{\lambda_i\}$ can be estimated by
maximum likelihood using standard software for generalized linear
models, with a suitably specified model matrix. The primary purpose of
the **BradleyTerry2** package [@turn:12], implemented in the
R statistical computing environment [@ihak:gent:96; @R], is to
facilitate the specification and fitting of such models and some
extensions.

The **BradleyTerry2** package supersedes the earlier **BradleyTerry**
package [@firt:05], providing a more flexible user interface to allow a
wider range of models to be fitted. In particular,
**BradleyTerry2** allows the inclusion of simple random effects so that
the ability parameters can be related to available explanatory variables
through a linear predictor of the form
$$\lambda_i=\sum_{r=1}^p\beta_rx_{ir} + U_i.$$
The inclusion of the prediction error $U_i$ allows for variability
between players with equal covariate values and induces correlation
between comparisons with a common player. **BradleyTerry2** also allows
for general contest-specific effects to be included in the model and
allows the logit link to be replaced, if required, by a different
symmetric link function (probit or cauchit).

The remainder of the paper is organised as follows.
Section [<span class="ref" data-target="sec:BTmodel"></span>](#sec:BTmodel) demonstrates how to use the **BradleyTerry2**
package to fit a standard (i.e., unstructured) Bradley-Terry model, with
a separate ability parameter estimated for each player, including the
use of bias-reduced estimation for such models.

# Acknowledgments {#acknowledgments .unnumbered}

This work was supported by the UK Engineering and Physical Sciences
Research Council.

# References {#references .unnumbered}
:::
