---
title: 'Exact Affine Stone Index Demand System in : The *easi* Package'
abstract: |
  *easi* is a package for that enables the estimation of the Exact
  Affine Stone Index (EASI) demand system proposed by
  [Pendakur]{.smallcaps} and [Lewbel]{.smallcaps}. The EASI system is
  more flexible and easier to manipulate than traditional demand systems
  such as AIDS or QAIDS. It offers four major advantages: First, EASI
  budget shares are linear in parameters, conditional on real
  expenditures. Second, EASI demands are not constrained by the
  theoretic rank limit of [Gorman]{.smallcaps}. Third, unobserved
  preferences heterogeneity are taken into account via EASI error terms,
  which are equivalent to random utility parameters. Finally, EASI
  demands can be polynomials or splines of any order. Estimation of the
  EASI demand system has already been implement in Stata by
  [@Pendakur/08]. The *easi* package is more than a simple port of the
  Stata code. It offers a number of extensions within an unified
  framework : calculation of elasticities and equivalent income,
  simulations, selection of interaction terms, graphical representations
  of Engel curves with/without confidence intervals, *etc.*
author: |
  Stéphane Hoareau\
  Université Laval Guy Lacroix\
  Université Laval Mirella Hoareau\
  Université Laval Luca Tiberti\
  Université Laval
date: '2024-07-08'
vignette: |-
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Exact Affine Stone Index Demand System inPackage}
  %\VignetteDepends{systemfit, stats, micEcon}
output:
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    number_sections: yes
bibliography: biblioeasi.bib

---

::: article
# Introduction

``` {r echo=FALSE,results='hide'}
options(prompt= "R> ", useFancyQuotes = FALSE)
```

Most empirical analyses of consumer expenditure data rely on parametric
demand models because they are relatively easy to estimate. Yet, it is
well-known that these models are plagued with a number of empirical and
theoretical shortcomings. Indeed, recent work has shown that many goods
depict Engle curves that are highly nonlinear, even S-shaped
([@Blundell/09]), a feature parametric models simply cannot account for.
Furthermore, all parametric models face the so-called Gorman-type rank
restrictons ([@Gorman/81]). Finally, unobserved heterogeneity of
preferences cannot readily be incorporated into most parametric models.
Traditionally, the error terms are treated as random utility parameters
([@Brown/89], [@McFadden/90], [@Brown/98], [@Lewbel/01], and
[@Beckert/04]).

The Exact Affine Stone Index (EASI) Demand System of [@Pendakur/08] and
[@Lewbel/09] is a major breakthrough in the estimation of demand
systems. It overcomes the aforementioned shortcomings while remaining
"easi" to use. The main novelty of their approach is to express utility
in terms of observed variables as derived from the Hicksian demand
functions. [Pendakur]{.smallcaps} and [Lewbel]{.smallcaps} thus suggest
we focus on what they refer to as *implicit* Marshallian demand
functions. The latter are simply Hicksian demand functions in which the
unobserved utility level is substituted out. Their *implicit* demand
system avoids all the caveats mentionned above: It is linear in
parameters, can easily incorporate unobserved heterogeneity, is not
limited by the Gorman rank restrictions, it is capable of generating
highly nonlinear Engle curves, and most of all, it is relatively "easi"
to estimate.

[Pendakur]{.smallcaps} and [Lewbel]{.smallcaps} derive the theoretical
properties of the EASI model and propose an estimation strategy
(iterated 3SLS). In addition, they derive various budget/quantity
elasticities. Finally, they illutrate the properties of the model using
Canadian micro-data.

Currently, offers two solutions for estimating demand systems. The
*systemfit* package ([@Hen/11]) can estimate systems of linear and
nonlinear equations using Ordinary Least Squares (OLS), Weighted Least
Squares (WLS), Seemingly Unrelated Regressions (SUR), Two-Stage Least
Squares (2SLS), Weighted Two-Stage Least Squares (W2SLS), and
Three-Stage Least Squares (3SLS). The *micEconAids* package
([@Hen/11bis]) focuses explicitely on the Almost Ideal Demand System
(AIDS) suggested by Deaton and Muellbauer (1980). It is also based upon
*systemfit*.

The *easi* package offers a unified framework within which the user can
effortlessly estimate the EASI demand system as well as request
additional statistical analyses. Thus in addition to the estimation of
the model, the *easi* package can calculate predicted budget shares,
generate graphical representations of Engel curves (with/without
confidence intervals), calculate various budget shares (price, income,
demographics) and quantity (price, income) elasticities, as well as
calculate equivalent incomes. The package also simulates price changes,
income changes, demographics changes and measures the impact on
predicted budget shares and elasticities and on the shapes of Engel
curves. The *easi* package is more flexible than the Stata code proposed
by [@Pendakur/08] in that the user can choose a subset of demographic
variables to interact with prices and/or expenditure.

The paper is organized as follows: Section [2](#sec:model) presents the
EASI Demand System. The EASI model, the associated calculations and
estimation method are presented in details. Section [3](#sec:modelelast)
focuses on the calculations of the elasticities. We briefly recall the
expressions for the budget shares presented in [Lewbel]{.smallcaps} and
[Pendakur]{.smallcaps} (2009), after which we derive similar expressions
for the quantity elasticities. Section  [4](#sec:equivincome) develops
the expressions for the equivalent incomes in the context of EASI model.
Section  [5](#sec:concavity) provides a test for the local concavity of
the EASI cost function. Section  [6](#sec:structure) presents the *easi*
package structure. Section  [7](#sec:examples) illustrate the use of
*easi* package with several examples. In particular, we replicate the
estimation results of [@Lewbel/09]. Section [8](#sec:conclusion)
concludes.

# The EASI Demand System {#sec:model}

Let $C(p,u,z,\epsilon)$ be a cost (expenditure) function, where $p$ is
the price vector, $u$ is the utility level, $z$ is the a vector of
demographic variables which proxy observable preference heterogeneity.
In additon, let $\varepsilon$ be a vector of error terms which include
unobservable preference heterogeneity. Hicksian compensated
budget-shares functions can be derived using Shephard's lemma:
$w=\omega(p,u,z,\varepsilon)=\nabla_p
C(p,u,z,\epsilon)$. By expressing the indirect utility function in terms
of $g$ of $w,p,x,z$, the *implicit* utility function, $y$, is defined as
$y=g(\omega(p,u,z,\varepsilon),p,x,z)=g(w,p,x,z)$. The *implicit*
utility function depends only on observable data. Its closed-form
expression is flexible, easily lends itself to empirical implementation,
and does not depend on the utility have itself a closed-form
expressions. The *implicit* Marshallian demand system is defined as
$w=\omega(p,y,z,\varepsilon)$, which is simply the Hicksian demand
system with $y$ substituted in for $u$.

[Lewbel]{.smallcaps} and [Pendakur]{.smallcaps} (2009) refer to this
class of cost functions as "Exact Affine Stone Index (EASI) cost
functions\", where $y$ corresponds to an affine function of the Stone
index deflated by the log nominal expenditures.

[Lewbel]{.smallcaps} and [Pendakur]{.smallcaps} propose the following
cost function, which is particularly convenient for empirical
implementation :

$$\label{eq1} \begin{array}{l} \ln C(p,y,z,\varepsilon)
  = {\displaystyle y + \sum_{j=1}^J m^j(y,z)\ln p^j
  +\frac{1}{2}\sum_{j=1}^J\sum_{k=1}^J a^{jk}(z) \ln p^j \ln p^k
  }\\[16pt] \phantom{\ln C(p,u,z,\varepsilon) = } {\displaystyle +
  \frac{1}{2}\sum_{j=1}^J\sum_{k=1}^J b^{jk} \ln p^j \ln p^k y +
  \sum_{j=1}^J \varepsilon_j \ln p^j} \end{array}   (\#eq:eq1)$$

Assuming there are $J$ goods and $T$ demographic variables, they propose
the following parameterisation :

$$\label{eq2}
 m^j(y,z) = \sum_{r=1}^R b_r^j y^r +\sum_{t=1}^T g_t^j z^t + \sum_{t=2}^T h_t^j z^t y   (\#eq:eq2)$$

and $a^{jk}(z)=\sum_{t=1}^T a^{jkt}z_t$

The *implicit* Marshalian budget shares for each $j \in 1...J$ is then
given by:

$$\label{eq5} w^j = \sum_{r=1}^R b_r^j y^r +
  \sum_{t=1}^T g_t^j z_t + \sum_{k=1}^T \sum_{t=1}^T a_{jkt} z_t \ln
  p^k + \sum_{k=1}^J b_{jk} \ln p^k y + \sum_{t=2}^T h_{t}^j z_t y +
  \varepsilon_j   (\#eq:eq5)$$

$$\label{eq6} y = \frac{\ln x -
\sum_{j=1}^J w_j \ln p^j+ 1/2 \sum_{j=1}^J \sum_{k=1}^J a_{jkt} z_t
\ln p^j \ln p^k}{1-\frac{1}{2}\sum_{j=1}^J \sum_{k=1}^J b_{jk} \ln p^j
\ln p^k}   (\#eq:eq6)$$

These *implicit* Marshalian budget shares have several desirable
properties. The linearity in parameters and additive error terms -
namely the random utility parameters $\varepsilon_j$ representing
unobserved preferences - are certainly two of them. In addition, price
effects can easily be interacted with expenditures and demographic
characteristics. Engel curves can take virtually any shape through
arbitrary high-order polynomials in log real expenditures. The
demographic variables enter both through the intercept and the slopes of
the log real-expenditures. Finally, EASI Engel curves for each good are
almost completely unrestricted.

The strict monotonicity and concavity of the cost function are required
in order to satisfy the others desirable properties of demand analysis -
regularity, adding-up constraints, homogeneity and Slutsky symmetry.

Three methods of estimation are proposed in [@Pendakur/08] and
[@Lewbel/09]. All three take into account the heteroskedasticity of
errors terms and the endogeneity of $y$. Parameters may be estimated by
using [Hansen]{.smallcaps}'s (1982) General Method of Moments or by
homoskedastic nonlinear 3SLS. The estimators are consistent with
heteroskedasticity but are only asymptotically efficient when the errors
terms are homoskedastic. The authors recommend using an iterative linear
3SLS, which is a special case of a fixed-point based estimator
considered by Dominitz and Sherman (2005).

# Calculation of the elasticities {#sec:modelelast}

## The Budget Shares Elasticities {#sec:modelelast1}

Five types of budget shares elasticities are calculated in
[@Pendakur/08] and [@Lewbel/09]:

The semi elasticities of budget shares, $\Psi$, are given by

:   :
    $$\label{eq7}
      \Psi = \frac{\partial \omega_j^i(p,y,z,\varepsilon)}{\partial \ln p^k} = \sum_{t=1}^T a^{jkt}z_t + b^{jk}y   (\#eq:eq7)$$

The real expenditure semi-elasticities, $\aleph$, are given by

:   :
    $$\label{eq8}
        \aleph = \frac{\partial \omega_j^i(p,y,z,\varepsilon)}{\partial y} = \sum_{r=1}^R b_r^{j}r y^{r-1} + \sum_{t=2}^T h_t^{j} z_t  + \sum_{k=1}^J b^{jk} \ln p^k   (\#eq:eq8)$$

The semi elasticities with respect to observable demographics, $\zeta$, are given by

:   :
    $$\label{eq9}
        \zeta = \frac{\partial \omega_j^i(p,y,z,\varepsilon)}{\partial z_t} =  g_t^{j} + h_t^{j} y  + \sum_{k=1}^J a^{jkt} \ln p^k   (\#eq:eq9)$$

The compensated quantity derivatives with respect to prices, $\Gamma$ ,are given by 

:   :
    $$\label{eq10}
      \Gamma = W^{-1}(\Psi+ww')  \textrm{ where }  W=diag(w)   (\#eq:eq10)$$

The compensated expenditures elasticities with respect to prices, $S$, are given by

:   :
    $$\label{eq11}
      S=\Psi+ww'-W   (\#eq:eq11)$$

## The Quantities Elasticities {#sec:modelelast2}

The elasticities for the quantities have been developed as part of the
construction of the *easi* package.

Consider the EASI implicit marshallian demand system (\@ref(eq:eq5) and
\@ref(eq:eq6)) and the following identity:

$$\label{eq12}
w^j = \frac{p_j Q_j}{x},   (\#eq:eq12)$$

where $p_j$ is the nominal price of good $j$, $Q_j$ is the amount of
good $j$ and $x$ is the total expenditure.

Elasticity of good $j$ with respect to the price of good $i$ is given
by:

$$\label{eq13}
\frac{\partial Q_j}{\partial p_i} \frac{p_i}{Q_j} = \eta_j^i   (\#eq:eq13)$$

Therefore:

$$\label{eq14}
\frac{\partial Q_j}{\partial p_i} = \frac{\partial \left(\frac{x w_j}{p_j}\right)}{\partial p_i} + \frac{x}{p_j} \frac{\partial w_j}{\partial p_i}   (\#eq:eq14)$$

Moreover :

$$\label{eq15}
\frac{\partial w_j}{\partial p_i} = A1 + A2 + A3 + A4,   (\#eq:eq15)$$

with:

$$\label{eq16}
A1 = \frac{\partial (\sum_{r=1}^R b_r^j y^r)}{\partial p_i}   (\#eq:eq16)$$

$$\label{eq17}
A2 = \frac{\partial (\sum_{t=2}^T h_t z_t y) }{\partial p_i}   (\#eq:eq17)$$

$$\label{eq18}
A3 = \frac{\partial (\sum_{k=1}^J \sum_{t=1}^T a_{jkt} z_t \ln p_k)}{\partial p_i}   (\#eq:eq18)$$

$$\label{eq19}
A4 = \frac{\partial (\sum_{k=1}^J b_{jk} \ln p_k y)}{\partial p_i}   (\#eq:eq19)$$

Calculations of $A_i$ give:

$$\label{eq20}
 A1 = \frac{\partial y}{\partial P_i} \sum_{r=1}^R b_r^j y^{r-1}   (\#eq:eq20)$$

$$\label{eq21}
A2 =  \frac{\partial y}{\partial P_i} \sum_{t=2}^T h_t z_t   (\#eq:eq21)$$

$$\label{eq22}
A3 =  \sum_{t=1}^T a_{jkt} z_t \frac{1}{p_i}   (\#eq:eq22)$$

$$\label{eq23}
A4 =  \frac{1}{p_i} b_{ji} y + \frac{\partial y}{\partial p_i} \sum_{k=1}^J b_{jk} \ln p_k   (\#eq:eq23)$$

Let $C = \sum_{r=1}^R b_r^j y^{r-1}$, $D =  \sum_{t=2}^T h_t z_t$,
$E = \sum_{t=1}^T a_{jkt} z_t$, $F = b_{ji} y$ and
$G = \sum_{k=1}^J b_{jk} \ln p_k$

Moreover

$$\label{eq23p}
y = \frac{u(p_i)}{v(p_i)},  \textrm{ where :}   (\#eq:eq23p)$$

$$\label{eq24}
u(p_i)= \ln x - \sum_{j=1}^J w_j \ln p^j+ 1/2 \sum_{j=1}^J \sum_{k=1}^J \sum_{t=1}^T a_{jkt} z_t \ln p^j \ln p^k   \textrm{ and }   (\#eq:eq24)$$

$$\label{eq24p}
v(p_i) = 1-\frac{1}{2}\sum_{j=1}^J \sum_{k=1}^J b_{jk} \ln p^j \ln p^k   (\#eq:eq24p)$$

It follows that:

$$\label{eq25}
 u'(p_i) = \frac{1}{P_i} ( - w_i + \sum_{k=1}^J \sum_{t=1}^T a_{jkt} z_t \ln p^k)   \textrm{ and }   (\#eq:eq25)$$

$$\label{eq25p}
 v'(p_i) = - \frac{1}{P_i} b_{jk} \ln p^k   (\#eq:eq25p)$$

Furthermore :

$$\label{eq26}
\frac{\partial y}{\partial P_i} = \frac{u'(p_i)v(p_i) -u(p_i)v'(p_i) }{v^2(p_i)}   (\#eq:eq26)$$

A little algebra allows us to write :

$$\label{eq27}
\frac{\partial y}{\partial p_i} =  \frac{1}{p_i} B   (\#eq:eq27)$$

Where

$$\label{eq28}
B =  p_i \frac{u'(p_i)v(p_i) -u(p_i)v'(p_i) }{v^2(p_i)}   (\#eq:eq28)$$

Hence

$$\label{eq29}
\frac{\partial Q_j}{\partial p_i} = \frac{\partial \left(\frac{x w_j}{p_j}\right)}{\partial p_i} + \frac{x}{p_j} \frac{1}{p_i} (B(C+D+G) + E + F)   (\#eq:eq29)$$

Let $H=B(C+D+G) + E + F$.

Moreover

$$\frac{\partial \left(\frac{x w_j}{p_j}\right)}{\partial p_i} =\left\{ \begin{array}{lll} &-\frac{x w_j}{p_j^2} &\textrm{if } i=j \\
 & 0 & \textrm{otherwise} \\
\end{array}\right.$$

$$\label{eq31}
\frac{\partial Q_j}{\partial p_i} = \frac{Q_j}{p_i} \left( -1 + \frac{H}{w_j}  \right)   (\#eq:eq31)$$

Hence, letting $\eta_j^i$ be the elasticity of good $j$ with respect to
the price of good $i$, we obtain :

$$\label{eq32}
 \eta_j^i = -1 (i==j) + \frac{H}{w_j}   (\#eq:eq32)$$

Consider the EASI implicit marshallian demand system (\@ref(eq:eq5),
\@ref(eq:eq6)) and the identity (\@ref(eq:eq12)):

Let $\eta_j^x$ be the elasticity of good $j$ with respect to income.
This elasticity is given by:

$$\label{eq33}
\frac{\partial Q_j}{\partial x} \frac{x}{Q_j} = \eta_j^x   (\#eq:eq33)$$

So, we have :
$$\label{eq34}
\frac{\partial Q_j}{\partial x} = \frac{w_j}{p_j} + \frac{x}{p_j} \frac{\partial w_j}{\partial x}   (\#eq:eq34)$$

Moreover :

$$\label{eq35}
 \frac{\partial w_j}{\partial x} = \frac{1}{x} (C+D+G)   (\#eq:eq35)$$

This allows us to write :

$$\label{eq36}
\frac{\partial Q_j}{\partial x} = \frac{1}{P_j} w_j + \frac{x}{p_j} \frac{1}{x} (C+D+G)   (\#eq:eq36)$$

Hence, we obtain :

$$\label{eq37}
 \eta_j^x = 1 + \frac{C+D+G}{w_j}   (\#eq:eq37)$$

where $C = \sum_{r=1}^R b_r^j y^{r-1}$, $D = \sum_{t=2}^T h_t z_t$ and
$G = \sum_{k=1}^J b_{jk} \ln p_k$.

# Calculation of the equivalent income {#sec:equivincome}

The equivalent income is defined as the income level, $e_{c,h}$, that
insures the utility levels are the same when evaluated at two prices
vectors, *i.e.*:

$$\label{eq38}
v (p_c,x_{c,h}) = v (p_r, e_{c,h}),   (\#eq:eq38)$$

where $v (.)$ is the indirect utility function, $p_r$ is the reference
price, and $p_c$ is a different price vector. By inverting the indirect
utility function, we obtain the equivalent income in terms of
expenditure function: $e_{c,h} = e (p_r, p_c,
x_{c,h})$, where $e_{c,h}$ is the equivalent income of household $h$
living in stratum c, facing the price vector $p_c$, with a level of
nominal income per capita (or per adult equivalent) $x_{c,h}$. The
equivalent income $e_{c,h}$ is the level of income, at the reference
price $p_r$, offers the same utility level than that obtained with the
income level $x_{c,h}$ and the price system $p_c$. The function
$e (p_r, p_c, x_{c,h})$ is increasing with respect to $p_r$ and
$x_{c,h}$, decreasing with $p_c$, concave and homogeneous of degree one
with respect to the reference price, and has continuous first and second
derivatives in all its arguments.

Consider the cost function (\@ref(eq:eq1)) in the EASI class, where $y$
is replaced by $u$ . From (\@ref(eq:eq2)) and (\@ref(eq:eq5)), we have :

$$\label{eq42}
 m^j(u,z) = w^j(p,u,z) - \sum_{k=1}^J a_{jk} \ln p^k - \sum_{k=1}^J b_{jk} \ln p^k u   (\#eq:eq42)$$

By substituting (\@ref(eq:eq42)) in (\@ref(eq:eq1)), we have :

$$\label{eq43} \begin{array}{l} \ln C(p,u, z) =
{\displaystyle u(1-\sum_{j=1}^J \sum_{k=1}^J b_{jk} \ln p^j \ln p^k) +
\sum_{j=1}^J \left(w^j(p,u,z) - \sum_{k=1}^J a_{jk} \ln p^k \right)
\ln p^j }\\[16pt] \phantom{\ln C(p,u, z) =} {\displaystyle +
\frac{1}{2} \sum_{j=1}^J \sum_{k=1}^J a_{jk} \ln p^j \ln p^k }
\end{array}   (\#eq:eq43)$$

The contemporary situation is characterized by nominal total
expenditures, $x_{c,h}$, and prices, $p_c$. This configuration achieves
a level of utility $\overline{u}$:

$$\label{eq44} \overline{u} = \frac{\ln x_{c,h} -
\sum_{j=1}^J w^j \ln p_c^j + \frac{1}{2} \sum_{j=1}^J \sum_{k=1}^J
a_{jk} \ln p_c^j \ln p_c^k}{1-\frac{1}{2} \sum_{j=1}^J \sum_{k=1}^J
b_{jk} \ln p_c^j \ln p_c^k}   (\#eq:eq44)$$

The reference or *ex ante* situation is characterized by nominal total
expenditures equal to the equivalent income, $e_{c,h}$, and prices,
$p_r$: This configuration also achieves a level of utility
$\overline{u}$. We can calculate this equivalent income $e_{c,h}$ by
solving :

$$\label{eq45} \begin{array}{l} \ln C(p_r,u, z) = \ln
  e_{c,h} ={\displaystyle \overline{u}(1-\frac{1}{2}\sum_{j=1}^J
  \sum_{k=1}^J b_{jk} \ln p_c^j \ln p_c^k) + \sum_{j=1}^J
  \left(w^j(p_r,u,z) - \sum_{k=1}^J a_{jk} \ln p_r^k \right) \ln p_r^j
  }\\[16pt] \phantom{\ln C(p_r,u, z) = \ln e_{c,h} =} {\displaystyle +
  \frac{1}{2} \sum_{j=1}^J \sum_{k=1}^J a_{jk} \ln p_r^j \ln p_r^k }
  \end{array}   (\#eq:eq45)$$

By substituting (\@ref(eq:eq44)) in (\@ref(eq:eq45)), we obtain:

$$\label{eq46} \begin{array}{l} e_{c,h} = \exp \Bigg(
  \ln x_{c,h} + \sum_{j=1}^J w^j \ln p_r^j - \sum_{j=1}^J w^j \ln
  p_c^j + \\ \phantom{e_{c,h} = \exp \Bigg( } + \frac{1}{2}
  \sum_{j=1}^J \sum_{k=1}^J a_{jk} \ln p_c^j \ln p_c^k - \frac{1}{2}
  \sum_{j=1}^J \sum_{k=1}^J a_{jk} \ln p_r^j \ln p_r^k \Bigg)
  \end{array}   (\#eq:eq46)$$

# Concavity of the cost function {#sec:concavity}

The EASI demand system estimation and the calculation of the equivalent
income assume that the cost function is concave. However, this concavity
can be checked after estimation. Furthermore, it is well known that a
semi-negative definite Hessian matrix is a necessary and sufficient
condition to consider that the cost function is concave.

The Hessian matrix, $H$, is defined as:

$$\label{eq100}
H=\frac{\partial^2 C}{\partial p_j \partial p_k}  \textrm{ j,k } \in [1...J]   (\#eq:eq100)$$

where $C$ is the EASI cost function described in (\@ref(eq:eq1)). By
naming $S$ the right hand side in the equation (\@ref(eq:eq1)), we can
write:

$$\label{eq101}
\ln C = S \Rightarrow C = \exp\{S\}   (\#eq:eq101)$$

The Hessian matrix, $H$, is also equal to:

$$\label{eq102}
H = \frac{\partial^2 S}{\partial p_j \partial p_k} \exp{S} + \frac{\partial S}{\partial p_j} \frac{\partial S}{\partial p_k} \exp{S}   (\#eq:eq102)$$

where:

$$\frac{\partial^2 S}{\partial p_j \partial p_k} = 2\frac{\partial^2 y}{\partial p_j \partial p_k} + \frac{\partial^2 S0}{\partial p_j \partial p_k} + \frac{\partial^2 S1}{\partial p_j \partial p_k}  + \frac{\partial^2 S2}{\partial p_j \partial p_k}y + \frac{\partial S2}{\partial p_j} \frac{\partial y}{\partial p_k}  + \frac{\partial S2}{\partial p_k} \frac{\partial y}{\partial p_j}$$

and

$$\frac{\partial S}{\partial p_j} = \frac{\partial y}{\partial p_j} + \frac{\partial S0}{\partial p_j} + \frac{\partial S1}{\partial p_j} + \frac{\partial S2}{\partial p_j} y + S2 \frac{\partial y}{\partial p_j}$$

A little algebra allows us to write:

$$S0 = \sum_{j=1}^J w^j \ln p_j$$

$$\frac{\partial S0}{\partial p_j} = \frac{w^j}{p_j}$$

$$\begin{array}{l}
 {\displaystyle \frac{\partial^2 S0}{\partial p_j \partial p_k} = 0 \textrm{ if } j \neq k } \\[12pt]
 \phantom{\frac{\partial^2 S0}{\partial p_j \partial p_k}} {\displaystyle = -\frac{w^j}{p_j^2} \textrm{ otherwise } }
 \end{array}$$

$$S1 = -1/2 \sum_{j=1}^J \sum_{k=1}^J \sum_{t=1}^T a_{jkt} z_t \ln p^j \ln p^k$$

$$\frac{\partial S1}{\partial p_j} = -\sum_{k=1}^J \sum_{t=1}^T \frac{a_{jkt}}{p_j} z_t \ln p^k$$

$$\frac{\partial^2 S1}{\partial p_j \partial p_k} = -\sum_{t=1}^T \frac{a_{jkt}}{p_jp_k}z_t$$

$$S2 = -1/2 \sum_{j=1}^J \sum_{k=1}^J  b_{jk} \ln p^j \ln p^k$$

$$\frac{\partial S2}{\partial p_j} = -\sum_{k=1}^J  \frac{b_{jk}}{p_j} \ln p^k$$

$$\frac{\partial^2 S2}{\partial p_j \partial p_k} = - \frac{b_{jk}}{p_jp_k}$$

# Package structure {#sec:structure}

![Figure 1: []{#fig:schemaeasi1 label="fig:schemaeasi1"} The Package
Structure.](schemaeasi.png){#fig:schemaeasi1 width="19cm" height="24cm"}

Figure [1](#fig:schemaeasi1) shows the *easi* package structure. The
package has five main functions. The first function, namely the `easi`
function, allows estimation of the model and generates nu,erous results
as an object of class `easi`. Five methods were written to retrieve the
results more easily: `coef` (parameter estimates), `predict` (matrix of
the fitted budget shares), `residuals` (matrix of residuals), `summary`
(summary of the estimation results), and `vcov` (covariance matrix of
the paramter estimates). The `easi` function uses the *systemfit*
package for the estimation procedure (iterated three stages least
squares).

The results of the `easi` function (coefficients, variance matrix) can
be used to compute Engel curves, elasticities, as well as to conduct
simulations. More precisely, the `engel` function computes and draws the
Engel Curves while the `elasticities` function calculates the
elasticities of the budget shares and the elasticities of the
quantities. The concavity of the cost function can be checked with the
`concavity` function.

Likewise, two types of simulations are implemented within the *easi*
package. Both require the user to specify new prices and / or new
demographics and / or a new vector of total expenditure. The
`simulations` generate new budget shares and `equiv.income` functions
and equivalent income after above mentioned changes.

The result of `simulations` function is also an object of class `easi`.
Therefore, the new elasticities and the new Engel Curves can be
calculated after the simulations.

An internal function allows the calculation of intermediate blocks
useful for generating the matrix of budget shares and the vector of
utility implied. This function, namely `intermediate.blocs` is called by
the `engel`, `elasticities`, `simulations` and `equiv.income` functions.

Finally, a database for examples and help is provided in the *easi*
package. These data are those used by [@Lewbel/09], namely the
`hixdata`.

# Examples {#sec:examples}

The *easi* package is loaded using:

``` {r }
library(easi)
```

To illustrate the use of the *easi* package, we use the `hixdata` data
frame (See [@Lewbel/09]).

``` {r }
data(hixdata)
```

Data consist of 4,847 observations of rental-tenure single-member
canadian households that had positive expenditures on rent, recreation,
and transportation (For details see [@Lewbel/09]).

The covariates of this data are:

`obs`:

:   number of observations

`sfoodh`:

:   the budget share of food at home

`sfoodr`:

:   the budget share of others foods

`srent`:

:   the budget share of rent

`soper`:

:   the budget share of household operations

`sfurn`:

:   the budget share of household furnishing and equipment

`scloth`:

:   the budget share of clothing

`stranop`:

:   the budget share of transportation operations

`srecr`:

:   the budget share of recreations

`spers`:

:   the budget share of personal care

`pfoodh`:

:   the logarithm of the price of food at home

`pfoodr`:

:   the logarithm of the price of others foods

`prent`:

:   the logarithm of the price of rent

`poper`:

:   the logarithm of the price of household operations

`pfurn`:

:   the logarithm of the price of household furnishing and equipment

`pcloth`:

:   the logarithm of the price of clothing

`ptranop`:

:   the logarithm of the price of transportation operations

`precr`:

:   the logarithm of the price of recreations

`ppers`:

:   the logarithm of the price of personal care

`log_y`:

:   the logarithm of total expenditure

`age`:

:   the person's age minus 40

`hsex`:

:   the sex dummy equal to one for men

`carown`:

:   a car-nonowner dummy equal to one if real gasoline expenditures (at
    1986 gasoline prices) are less than 50 dollars

`time`:

:   a time variable equal to the calendar year minus 1986

`tran`:

:   a social assistance dummy equal to one of government transfers are
    greater than 10 percent of gross income

`wgt`:

:   weighting variable

## Estimation

The estimation is performed using the `easi` function, which has the
following obligatory arguments: `shares` is the budget shares matrix
(one observation by row and one item by column), `log.price` is the
prices matrix (in logaritms) (one observation by row and one item by
column), `var.soc` is the matrix of demographic variables and `log.exp`
is the logaritm of total expenditure.

The user can customize the estimation with the following options. The
first option - `y.power` - is an integer which corresponds to the
highest desired power of `y` (implicit utility) in the system.
`labels.share` and `labels.soc` are two strings which contain
respectively the names of budget shares and the names of demographic
variables. The following options - `py.inter`, `zy.inter` and
`pz.inter` - are three logical variables which are each fixed to TRUE
(FALSE otherwise) if the user wants respectively to enable the
interaction between the price variables and `y`, the interaction between
the demographic variables and `y`, the interaction between the prices
and demographic variables. Finally, `interpz` is a vector which allows
to choose which demographic variables have to be crossed with the price.
For example, `interpz=c(3)` means that prices are crossed with the third
demographic variable while `interpz = c (1:n)` means that prices are
crossed with the first `n` demographic variables.

To illustrate the use of *easi* package, we reproduce the results of
[@Lewbel/09]. We first estimate a simple EASI model, which we
arbitrarily call `est`.

``` {r }
#****** Budget Shares Matrix *************
#     shares_HIX=hixdata[,2:10]
#****** Price Matrix (in logarithms) *****
#     log.price_HIX=hixdata[,11:19] 
#****** Demographic matrix **********
#     var.soc_HIX=hixdata[,21:25]
#****** Logarithm of total expenditure ***
#****** (here divised by a price index) **
#     log.exp_HIX=hixdata[,20]  
#****** Labels of demographic variables **
#     labels.soc <- c("age","hsex","carown","time","tran")
#****** Labels of budget shares **********
#     labels.share=c("food in","food out","rent","operations",
#       "furnishing","clothes","transport","recreation")
```

``` {r }
#est <- easi(shares=shares_HIX,log.price=log.price_HIX,var.soc=var.soc_HIX,
 #           y.power=5,log.exp=log.exp_HIX,labels.share=labels.share,
  #          labels.soc=labels.soc,py.inter=TRUE, zy.inter=TRUE, 
   #         pz.inter=TRUE, interpz=c(1:ncol(var.soc_HIX)))
```

Several methods for an easier manipulation of the results. The `coef`
method allows to recover the coefficients. Here, for more readability,
we present only the first three lines of the coefficient matrix.

``` {r }
#head(coef(est),3)
```

The `vcov` method allows to recover the variance matrix. The covariance
matrix is too large for printing here. Its size is given by:

``` {r }
#dim(vcov(est))
```

The `predict` method allows to recover the fitted budget shares:

``` {r }
#head(predict(est),3)
```

The `residuals` method allows to recover the residuals.

``` {r }
#head(residuals(est),3)
```

The `concavity` function allows to check the local concavity of the cost
function.

``` {r }
#head(concavity(est))
```

Here, the cost function is concave on more than 90$\%$ of the
observations.

## Elasticities

The elasticities can be calculated with the `elasticities` function. The
arguments of the function are the result of the `easi` function (an
object of class `easi`, here `est`), the type of desired elasticities
(between \"price\", \"demographic\" and \"income\") and a logical
variable `sd` that indicates if standard deviations must be calculated.

The calculation of price elasticities are performed by:

``` {r }
#elastprice <- elastic(est,type="price",sd=TRUE)
```

The price elasticities of budget shares are recovered by:

``` {r }
#elastprice$EP[paste("p",labels.share,sep=""),
 #             paste("p",labels.share,sep="")]
```

The corresponding standard deviations are recovered by:

``` {r }
#elastprice$EP_SE[paste("p",labels.share,sep=""),
  #               paste("p",labels.share,sep="")]
```

The elasticities of quantities with respect to prices are recovered by:

``` {r }
#elastprice$ELASTPRICE[paste("p",labels.share,sep=""),
#                      paste("p",labels.share,sep="")]
```

The corresponding standard deviations are recovered by:

``` {r }
#elastprice$ELASTPRICE_SE[paste("p",labels.share,sep=""),
#                         paste("p",labels.share,sep="")]
```

The calculation of income elasticities are performed by:

``` {r }
#elastincome <- elastic(est,type="income",sd=FALSE)
```

The income elasticities of budget shares are recovered by:

``` {r }
#elastincome$ER[1,labels.share]
```

The elasticities of quantities with respect to income are recovered by:

``` {r }
#elastincome$ELASTINCOME[1,labels.share]
```

The calculation of demographic elasticities are performed by:

``` {r }
#elastdemographic <- elastic(est,type="demographics",sd=FALSE)
```

## Engel curves

If one wants to calculate Engel curves, one can use the `engel`
function, whose arguments are the result of the `easi` function (an
object of class `easi`, here `est`), the name where the Engel curves
must be graphically represented, and a logical variable `sd` that
indicates if confidence intervals must be calculated and represented.

``` {r }
#eng1 <- engel(est,file="graph_engels_curves",sd=TRUE)
```

The figure [2](#fig:eng1) shows the Engel curves of the `est`
estimation. Each green circle is the median of the budget share for the
considered percentile of total expenditure described in abscissa.
Magenta crosses delimit a confidence interval of 95$\%$. Curves (black,
blue and red) correspond to three increasing levels of smoothing.

![Figure 2: []{#fig:eng1 label="fig:eng1"} The Engel Curves of easi
estimation.](grapheng1.png){#fig:eng1 width="17cm" height="17cm"}

## Simulations

*easi* provides routines to perform simulations. To illustrate, we
decide to reproduce the simulation of [@Lewbel/09], namely the estimated
Engel curves from the model for a 40-year-old car-owning female in 1986
who didn't receive much government transfer income and having
$\varepsilon=0$.

This configuration implies that all prices (in logarithms) and all
demographic variables are null.

``` {r }
 #    log.price_HIX.SIM1 <- log.price_HIX
 #   for (i in 1:ncol(log.price_HIX))
 #  log.price_HIX.SIM1[,i] <- 0
     
 #    var.soc_HIX.SIM1 <- var.soc_HIX
 #    for (i in 1:ncol(var.soc_HIX))
 #    var.soc_HIX.SIM1[,i] <- 0
```

The `simulations` function allows to calculate the fitted values of
budget shares after the previous changes in prices and demographics.

``` {r }
  #   sim <- simulations(est,log.price_new=log.price_HIX.SIM1,
  #           var.soc_new=var.soc_HIX.SIM1,log.exp_new=log.exp_HIX)
```

The corresponding Engel curves are calculated as previously.

``` {r }
#eng2 <- engel(sim,file="simeng",sd=TRUE)
```

The figure [3](#fig:simeng1) shows the Engel curves of the `sim`
simulation.

![Figure 3: []{#fig:simeng1 label="fig:simeng1"} The Engel Curves of
easi simulation.](simeng.png){#fig:simeng1 width="17cm" height="17cm"}

## Equivalent income

Our version of `hixdata` does not contain the variable \"total
expenditure\". We propose to simulate instead an \"hybrid model\" in
order to illustrate the use of the `equiv.income` function:

This model is composed of five budget shares whose means are
respectively equal to 0.25, 0.15, 0.20, 0.30 and 0.10.

``` {r }
w1 <- rnorm(3000,mean=0.25,sd=0.05)
w2 <- rnorm(3000,mean=0.15,sd=0.05)
w3 <- rnorm(3000,mean=0.20,sd=0.05)
w4 <- rnorm(3000,mean=0.30,sd=0.05)
w5 <- 1-w1-w2-w3-w4 
shares_SIM <- data.frame(w1,w2,w3,w4,w5)
```

We simulate five price vectors, whose means are respectively equal to
25, 15, 20, 30 and 10:

``` {r }
p1 <- log(rnorm(3000,mean=25,sd=3))
p2 <- log(rnorm(3000,mean=15,sd=2))
p3 <- log(rnorm(3000,mean=20,sd=3))
p4 <- log(rnorm(3000,mean=30,sd=4))
p5 <- log(rnorm(3000,mean=10,sd=1))
log.price_SIM <- data.frame(p1,p2,p3,p4,p5)
```

We simulate four demographics variable : V1, V3, V4 that are dummy
variables, and V2 that take his values in $\mathbb{N^+}$.

``` {r }
V1 <- abs(round(rnorm(3000,mean=0.7,sd=0.2)))
V2 <- abs(round(rnorm(3000,mean=2,sd=1)))+1
V3 <- abs(round(rnorm(3000,mean=0.7,sd=0.2)))
V4 <- abs(round(rnorm(3000,mean=0.7,sd=0.2)))
var.soc_SIM <- data.frame(V1,V2,V3,V4)
```

Finally, we simulate a vector of total expenditure whose the average is
1200.

``` {r }
     log.exp_SIM <- log(rnorm(3000,mean=1200,sd=200))
```

The first step is to estimate the EASI model:

``` {r }
     est2 <- easi(shares=shares_SIM,log.price=log.price_SIM,
                 var.soc=var.soc_SIM,log.exp=log.exp_SIM)
```

Let's consider the calculation of equivalent income after following
changes : PRICE_SIM are multiplied by 1.4 between reference and current
period while exp_SIM is only multiplied by 1.05 between reference and
current period. This scenario thus corresponds a priori to a loss of
purchasing power.

``` {r }
     log.price_SIM2 <- log(exp(log.price_SIM)*1.4)
     log.exp_SIM2 <- log(exp(log.exp_SIM)*1.05)
     equiv <- equiv.income(est2,log.exp_ref=log.exp_SIM,log.exp_cur=log.exp_SIM2,
                 log.price_ref=log.price_SIM,log.price_cur=log.price_SIM2)
```

The result indicates, as expected, that the welfare of households
decreased between the reference period and current period.

# Conclusion {#sec:conclusion}

*easi* aims at providing a unified framework allowing to estimate and
exploit the Exact Affine Stone Index (EASI) demand system of
[@Lewbel/09] not currently implemented in R. The EASI demand system has
several advantages in comparison to AIDS and QUAIDS. Firstly, its
numerical implementation is easier due to the linearity in terms of
parameters. Secondly, unobserved preferences are taken into account
thanks to additive errors which are interpreted as random utility
parameters. Finally, easi Engel Curves may have more complicated shapes
as suggested by the possible specification of high order polynomials in
log real expenditure in the system.

For each estimation, *easi* allows the choice of one version of EASI
model with one, two or all of the following interactions : interactions
between prices and log real expenditure, interactions between prices and
demographic variables and interactions between log real expenditure and
demographic variables.

It moreover offers tools for exploitation of the results and
simulations.

Among these tools, *easi* provides methods to retrieve more easily the
estimates, the residual, the variance matrix, the fitted budget shares
and the summary of estimation. It also develops functions to calculate
and draw the Engel Curves and functions to calculate elasticities (price
elasticities, income elasticities and demographic elasticities).

Furthermore, it enables simulations, namely the assessment of the impact
of price changes, income changes and demographics changes on fitted
budget shares and elasticities. Similarly, function for calculation of
equivalent income is available in *easi*.

Still, extensions and improvements of the software are under way,
notably the inclusion of weights.

Research is continuing in this direction.
:::
