---
title: A Simple Example for rnw_to_rmd Function
abstract: |
  Many statistics methods require one or more least squares problems to
  be solved. There are several ways to perform this calculation, using
  objects from the base R system and using objects in the classes
  defined in the `Matrix` package.

  We compare the speed of some of these methods on a very small example
  and on a example for which the model matrix is large and sparse.
author: |
  Yinxiang Huang\
  Beihang University\
  [21375146@buaa.edu.cn](21375146@buaa.edu.cn){.uri}
date: '2024-07-01'
vignette: |-
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{example-knitr.Rnw}
  %\VignetteDepends{}
output:
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    number_sections: yes
bibliography: ~

---

::: article
Sweave options will be automatically converted to knitr.

``` {r include=FALSE}
library(knitr)
opts_chunk$set(
tidy=FALSE, fig.keep='none'
)
```

To remove warning messages, setup code chunk use knitr option
`warning=FALSE`.

``` {r setup,include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
```

When the option OutDec is not `.`, put numbers in `\text`. See #348.

``` r
try this:
`r 1+1`
```

``` {r }
options(OutDec = ',')
```

This is the first test. abc `r 0.6` def

another test $a = `r 0.6`$.

and the last one $a = `r '0.6'`$.

``` {r }
options(OutDec = '.')
```

This is the first test. abc `r 0.6` def

another test $a = `r 0.6`$.

and the last one $a = `r '0.6'`$.

# Linear least squares calculations {#sec:LeastSquares}

``` {r test}
# Define the cars vector with 5 values
cars <- c(1, 3, 6, 4, 9)
plot(cars, type="o", col="blue")
title(main="Autos", col.main="red", font.main=4)
```

Let's test another image from R code chunk:

``` {r image}
library(Matrix)
data(CAex, package = "Matrix")
print(image(CAex, main = "image(CAex)")) # print(.) needed for Sweave
```
:::
